---
title: 'Data 621 Group 2 HW 3: Crime'
author: 'Members: Omar	Pineda,	Jeff	Littlejohn,	Sergio	Ortega	Cruz,	Chester	Poon,	Simon	Ustoyev'
date: 'Due: October 30, 2019'
output:
  html_document: 
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE)
```


## Assignment 


Build a binary logistic regression model on the training data set to predict whether the neighborhood will be at risk for high crime levels. You will provide classifications and probabilities for the evaluation data set using your binary logistic regression model. You can only use the variables given to you (or variables that you derive from the variables provided). Use 0.5 threshold. Below is a short description of the variables of interest in the data set: 
 
* zn: proportion of residential land zoned for large lots (over 25000 square feet) (predictor variable)
* indus: proportion of non-retail business acres per suburb (predictor variable)
* chas: a dummy var. for whether the suburb borders the Charles River (1) or not (0) (predictor variable)
* nox: nitrogen oxides concentration (parts per 10 million) (predictor variable)
* rm: average number of rooms per dwelling (predictor variable)
* age: proportion of owner-occupied units built prior to 1940 (predictor variable)
* dis: weighted mean of distances to five Boston employment centers (predictor variable)
* rad: index of accessibility to radial highways (predictor variable)
* tax: full-value property-tax rate per $10,000 (predictor variable)
* ptratio: pupil-teacher ratio by town (predictor variable)
* black: 1000(Bk - 0.63)2 where Bk is the proportion of blacks by town (predictor variable)
* lstat: lower status of the population (percent) (predictor variable)
* medv: median value of owner-occupied homes in $1000s (predictor variable)
* target: whether the crime rate is above the median crime rate (1) or not (0) **(response variable)**



## Write Up:
 
 
###1. Data Exploration

The dataset includes information on 466 neighborhoods in the city of Boston. Despite its East Coast location and reputation as a bastion of liberalism, Boston is among the most racially segregated of American cities. Attempts to integrate the schools using busing in the 1970s led to [sustained violence](https://en.wikipedia.org/wiki/Boston_desegregation_busing_crisis), including deaths. Recent scholarship has highlighted the widespread use of redlining, a process by which institutions such as banks refused to offer mortgages or other financial services to people of certain races if they wished to purchase a home in certain neighborhoods despite creditworthiness. 

In short, one would probably not want to construct a model to predict crime by neighborhood that uses variables such as race without having a clear idea of the model's intended use and an ethical framework for evaluating said model. This, however, is an academic exercise, so we proceed. Let's preview the data.

```{r,message=FALSE}
#load packages
library(knitr)
library(dplyr)
library(kableExtra)
```


```{r,message=FALSE}
#load training data
url_crime_train <- 'https://raw.githubusercontent.com/littlejohnjeff/Data_621_Fall_2019/master/crime-training-data_modified.csv'
crime_train_data <- read.csv(url_crime_train, header = TRUE)
kable(crime_train_data[1:15,]) %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

Expected variables are present. Note that, as indicated in the variables' descriptions, many of the variables have already been scaled or transformed in some way. Let's calculate summary statistics and generate a box plot for further review.

```{r,message=FALSE}
#summarize training data
library(psych)
kable(psych::describe(crime_train_data)) %>% kable_styling(bootstrap_options = c("striped", "hover", "responsive"))
```


We see 466 records in our training set and no missing values for any variable. Other than that, without a specific question in mind, it's difficult to draw any conclusions from this big table of numbers. We see no missing values that would require imputation using medians or other methods.

Now, we visualize using box plots. We'll separate the box plots by the target value, which signifies whether or not the neighborhood is high crime. And we'll approximate Boston Red Sox colors.

```{r,message=FALSE}
#boxplots of each variable split by target value
library(reshape2)
library(dplyr)
library(ggplot2)
crime_plot <- melt(crime_train_data, id.vars= 'target') %>% mutate(target = as.factor(target))

ggplot(data = crime_plot, aes(x = variable, y = value)) + geom_boxplot(aes(fill = target)) + facet_wrap( ~ variable, , dir = "h", scales = 'free') + scale_fill_manual(values=c("blue3", "red3"))
```

The dummy variable (chas) that represents proximity to the Charles River is not meaningful, but clear distinctions in distributions between the neighborhoods in which the crime rate is below and above the median - the target variable by which the box plots are split. We might later look at these values after transformations such as logs.

To check for skewness, let's examine the distribution of each variable independent of target variable value.

```{r,message=FALSE}
#histograms of training set
ggplot(crime_plot,aes(value)) + geom_histogram(bin=25,fill="Red3") + facet_wrap( ~ variable, , dir = "h", scales = 'free')
```

Skewness abounds. We will file this away for now and revisit in the Data Preparation part of the project. In particular, zn, nox, age, dis, ptratio, and lstat seem likely candidates for transformations.

We will now check for covariance.

```{r,message=FALSE}
#correlations
library(stats)
library(corrplot)
cor_train <- cor(crime_train_data, method="pearson") 
kable(cor_train, "html") %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

We see some very high positive and negative correlations between variables. Let's construct a more effective visualization.

```{r,message=FALSE}
#plot correlations
corrplot(cor_train)
```

We see candidates for combination due to covariance.

As a final step, let's look just a correlation between the independent variables and the target variables.

```{r}
#correlations just with target value
cor_train_target <- as.data.frame(cor_train) %>% dplyr::select(target)
kable(cor_train_target, "html") %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

Nox, or the concentration of nitrogen oxide, a significant pollutant that's harmful to human health, in a neighborhood, shows the closest correlation with the target variable at .73. Next, age, rad, tax, and indus all correlate with the target value just above .6. Zn showed the largest negative correlation with the target at -.43. Zn represents the percentage of residential lots zoned for large lots, which may be an indicator large rental housing - apartments.

  
  
 
###2. Data Preparation
 
Prior to modelling the training data set, we must prepare the data. We do not have any missing values, so imputation is not required. We will probably actually start with a model that uses all variables regardless of skew or covariance. However, we will definitely progress to using transformations and will also combine variables due to covariance in seeking the construction of accurate and valid models.

Let's look at how transformations might solve distribution issues with some of our variables. Earlier, we saw a strong right skew in the distribution of the variable lstat, which tracks the "lower status" of a neighborhood's population. Probably not the best phrasing.

```{r,message=FALSE}
#lstat distribution
ggplot(crime_train_data,aes(x=lstat)) + geom_histogram(bin=25,fill="Red3") 
```

What would a log transformation do to this distribution?

```{r,message=FALSE}
#lstat distribution after log transformation
ggplot(crime_train_data,aes(x=log(lstat))) + geom_histogram(bin=25,fill="Blue3") 
```

Looks slightly better. Let's generate log transformations for all variables in the dataset.

```{r,message=FALSE}
#histrogram with log transformations of all variables
ggplot(crime_plot,aes(log(value))) + geom_histogram(bin=25,fill="Blue3") + facet_wrap( ~ variable, , dir = "h", scales = 'free')
```

Medv looks slightly better. However, age remains strongly left skewed. Dis is now bimodal.

What about other transformations such as quadratic ones?

```{r,message=FALSE}
ggplot(crime_plot,aes(sqrt(value))) + geom_histogram(bin=25,fill="Yellow3") + facet_wrap( ~ variable, , dir = "h", scales = 'free')
```

Nope. Not a lot of improvement.

In Part 1, we saw high covariances among variables such as rad and tax (.91). To build the best models, we'll likely want to examine combining some of these variables that are correlated to each other, which tends to increase standard errors. This can lead to overfitting and inefficient models. We will not combine variables here but instead revisit this concept in part 3 when evaluating our models.

Our textbooks have also discussed the possibility of creating bins for continuous variables. For example, dis, the weighted distance of means of distances from a neighborhood to five Boston job centers, might be better suited to fall into three categories than to remain a continuous variable for performance reasons. 

Prior to building the model, we're going to split our training data into a true training set and a validation set. We'll go 80/20 training to validation.
```{r,warning=FALSE}
#split training data into true training and validation/tuning
train_size <- floor(0.8 * nrow(crime_train_data))

set.seed(123)
train_ind <- sample(seq_len(nrow(crime_train_data)), size = train_size)

crime_train_data_train <- crime_train_data[train_ind, ]
crime_train_data_valid <- crime_train_data[-train_ind, ]
```


###3. Build Models

Following convention, we will start with a model consisting of all variables, none of which have been transformed. While we've moved on to Part 3, where we will construct the models, the boundary between data preparation and model building is grey. We will to explore transformations and collinearity.

####Model 1 - All Variables Untransformed

```{r,message=FALSE}
#build model 1 - all variables untransformed
model1_untransformed <- glm(formula = target ~ ., family = "binomial", data = crime_train_data_train)
summary(model1_untransformed)
```

Our most significant variables generally tie to the variables we saw have the highest correlations with the target value earlier. We have an AIC of 169.67 and a residual deviance of 143.67. 

Let's run further diagnostics on the model. We will set a probability of .5 as being the cutoff for determining if a neighborhood will be high crime. Here, we check the relationship between the logit of the outcome and each predictive variable. (Target and the binary dummy variable chas should be ignored.) Again, these steps also could be labelled as data preparation.

```{r,message=FALSE}
#model 1 logit relationships
library(tidyr)
probabilities <- predict(model1_untransformed, type = "response")
predicted.classes <- ifelse(probabilities > 0.5, "pos", "neg")
head(predicted.classes)

mydata <- crime_train_data_train %>%
  dplyr::select_if(is.numeric) 
predictors <- colnames(mydata)

mydata <- mydata %>%
  mutate(logit = log(probabilities/(1-probabilities))) %>%
  gather(key = "predictors", value = "predictor.value", -logit)

ggplot(mydata, aes(logit, predictor.value))+
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  theme_bw() + 
  facet_wrap(~predictors, scales = "free_y")
```

Tax and zn do not show linear associations with the outcomes in logit scale. Along with the previously discussed lstat, they might benefit from transformations.

Let's use Cook's Distance to check for outliers.

```{r,message=FALSE}
#Cook's distance
#https://stat.ethz.ch/R-manual/R-devel/library/stats/html/plot.lm.html
plot(model1_untransformed, which = 4, id.n = 3)
```


```{r,message=FALSE}
#influential outlier checks
library(broom)
model1_untransformed.data <- augment(model1_untransformed) %>% 
  mutate(index = 1:n()) 

kable(model1_untransformed.data %>% top_n(3, .cooksd))  %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

An outlier is not necessarily influential. Let's check for that. 

```{r,message=FALSE}
#graph standardized residuals
ggplot(model1_untransformed.data, aes(index, .std.resid)) + 
  geom_point(alpha = .5) +
  theme_light()
```

Let's pull that point that's above 3 standardized residuals from 0.

```{r,message=FALSE}
kable(model1_untransformed.data %>% 
  filter(abs(.std.resid) > 3)) %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

Observation 338 is an influential outlier. 

Next, we check multicollinearity.

```{r,message=FALSE}
#model 1 vif
library(car)
car::vif(model1_untransformed)
```

The rule of thumb is that vif scores above 5 should be judged as having a high amount of multicollinearity. So rm and medv have issues in this regard.

In summary, we have:

1. Multiple predictors that do not have linear relationships with the logit of the outcome variable.
2. One influential outlier - index 338.
3. Two predictors with potentially problematically high multicollinearity.

The above are among many methods to check assumptions and diagnostics of logistic regression models. We will not repeat these steps - other than the summary diagnostics - for our additional attempts at constructing a model to predict high-crime neighborhoods.


####Model 2 - Collinearity

```{r,message=FALSE}
#build model 2
model2_coll <- glm(formula = target ~ zn + indus + chas + nox + age + dis + rad + tax + ptratio + lstat + medv, family = "binomial", data = crime_train_data_train)
summary(model2_coll)
```

Iniitally for model 2, we removed the variable with the highest collinearity (medv), which led to slightly higher residual deviance and AIC. Medv has a low p-value in the original model, so we instead removed rm, which led to a slight increase in residual deviance but a drop in AIC. Not a big difference.

####Model 3 - Partial Log Transformations

Here, we try log transformations of a couple variables that did not show linear relations between them and the logit of the outcome.

```{r,message=FALSE}
#build model 3
model3_some_logs <- glm(formula = target ~ zn + indus + chas + nox + rm + age + dis + rad + tax + ptratio + log(lstat) + log(medv), family = "binomial", data = crime_train_data_train)
summary(model3_some_logs)
```

Adding log transformations to lstat and medv has increased both our residual deviance and AIC, which is less than desirable.

####Model 4 - Significant Variables

Here, we include only the variables that were signfiicant (p value < .05) fro m model 1.

```{r,message=FALSE}
#build model 4
model4_sig_var <- glm(formula = target ~ nox + rad + dis + ptratio + medv + tax + age, family = "binomial", data = crime_train_data_train)
summary(model4_sig_var)
```

The residual deviance is higher here than in model 1 without a significant drop in AIC.

####Model 5 - Using LEAP for forward selection

We're not making good progress here on improving upon model 1, the initial try that included all untransformed variabes. Let's try the Leaps package, which will start with the most significant variable and then continue adding variables/features until many possible models have been evaluated. This is known as Forward selection. Note that while the Leaps pacage executed perfectly in R Studio and allowed us to evaluate numerous possible models  with the variables, it caused an error when "knitting."

The LEAPS package provided adjusted R^2 scores for a variety of packages. It started with just the nox variable and then added variables in constructing 12 total models. The 10h model (zn + indus + nox + rm + age+ dis + rad + tax + lstat + medv) from LEAPS has the highest R^2 but the 6th model is only slightly lower in adjusted R^2 with only six variables. In the interest of simplicity, we'll take that 6th model rom LEAPS - i-n-a-rd-t-m, which translates to the below.

```{r,message=FALSE}
#build model 5
model5_leap <- glm(formula = target ~ nox + rad + tax + medv + age + indus, family = "binomial", data = crime_train_data_train)
summary(model5_leap)
```

Note that we selected this model from LEAPS based on R^2 but are now comparing it to other models using residual deviance and AIC, where it is found wanting.  Deciding whether to use residual deviance, AIC, or adjusted R^2 among other model diagnostics like AICc and BIC is always challenge and may be domain - or maybe even problem - dependent. 

####Model 6 - Using stepAIC for forward selection

We now looked at using the stepAIC function and forward selection. As with the leaps package, we had problems "knitting" the stepAIC function from the MASS package here. Processed fine in R Studio, and led to the four-variable described below in model 6.

```{r,message=FALSE}
#build model 6
model6_step <- glm(formula = target ~ nox + rad + medv + age, family = "binomial", data = crime_train_data_train)
summary(model6_step)
```

In summary, for Part 3, Build Models, we attempted a number of methods to construct a good model. After setting a baseline with all variables, we attempted to address collinearity by removing a variable with a high vif score in Model 2. In Model 3, we applied logarithmic transformations to two of our variables the showd skewed distrubtions. In Model 4, we only included variables that showed significance in Model 1. Models 5 and 6 both used forward selection using R functions. 

Some of our findings were counterintuitive in that models with more variables showed higher AICs. Also, transfomrations did not seem effective, at least in terms of AIC and residual deviances.
 
###4. Select Models 

We will now run that remaining 20% of our training data through the models created with the 80% training set. We'll evaluate models 1, 2, 5, and 6. Finally, we will run the evaluation file that does not include target data through our model to ouptut predictions.

Model 1 confusion matrix and AUC information:
```{r,message=FALSE}
#model 1 confusion matrix
library(caret)
model1_predict <- predict(model1_untransformed, newdata=crime_train_data_valid, type="response")
model1_predict_data <- ifelse(model1_predict > .5, 1, 0)
confusionMatrix(data=model1_predict_data, crime_train_data_valid$target, positive='1')
```

```{r,message=FALSE}
#model 1 ROC
library('pROC')
plot(roc(crime_train_data_valid$target, model1_predict))
```

Now the same for model 2: 
```{r,message=FALSE}
#model 2 confusion matrix
model2_predict <- predict(model2_coll, newdata=crime_train_data_valid, type="response")
model2_predict_data <- ifelse(model2_predict > .5, 1, 0)
confusionMatrix(data=model2_predict_data, crime_train_data_valid$target, positive='1')
```

```{r,message=FALSE}
#model 2 roc
plot(roc(crime_train_data_valid$target, model2_predict))
```


And model 5:
```{r,message=FALSE}
#model 5 confusion matrix
model5_predict <- predict(model5_leap, newdata=crime_train_data_valid, type="response")
model5_predict_data <- ifelse(model5_predict > .5, 1, 0)
confusionMatrix(data=model5_predict_data, crime_train_data_valid$target, positive='1')
```

```{r,message=FALSE}
#model 5 roc
plot(roc(crime_train_data_valid$target, model5_predict))
```
 

Finally, model 6:
```{r}
#model 6 confusion matrix
model6_predict <- predict(model6_step, newdata=crime_train_data_valid, type="response")
model6_predict_data <- ifelse(model6_predict > .5, 1, 0)
confusionMatrix(data=model6_predict_data, crime_train_data_valid$target, positive='1')
```

```{r,message=FALSE}
#model 6 roc
plot(roc(crime_train_data_valid$target, model6_predict))
```
 
Models 1 and 2 provide better accuracy and AIC. Due to a slight edge in simplicity with having one fewer variable, we will go with model 2, which is just all untransformed variables other than the multicollinear rm. In reality, for a given business problem, there are often considerations at play that would lead one to value specificitiy over sensitivity, for example. Predicting that a neighborhood is high crime could be an expensive mistake, or it could be the safer assumption of the two, depending on context. Here, we don't know.
 
 Finally, we run the evaluation dataset through model 2, which is submitted along with this report. A reflection of the predictions of the evaluation set:
 
```{r,warning=FALSE}
#model 2 eval data histogram
url_crime_eval <- 'https://raw.githubusercontent.com/littlejohnjeff/Data_621_Fall_2019/master/crime-evaluation-data_modified.csv'
crime_eval_data <- read.csv(url_crime_eval, header = TRUE)
model2_eval_data <- predict(model2_coll, newdata=crime_eval_data, type="response")
model2_eval <- ifelse(model2_eval_data > .5, 1, 0)
hist(model2_eval)
crime_eval_data_out <- cbind(crime_eval_data,model2_eval)
```
 
 The above looks like it predicts neighborhoods to be high-crime at a slightly lower rate than indicated by the frequency distribution of the target value in the original full training set. We would need to know more about the sampling techinque used to split training from evaluation data before determining if this is problematic.
```{r}
#training data histogram
hist(crime_train_data$target)
```
 
 
```{r,warning=FALSE}
#write eval result file from model 2
write.csv(crime_eval_data_out,"DATA 621Data_621_Hw_3_Evaluation_Output.csv",row.names = FALSE)
```
 
<!-- Decide on the criteria for selecting the best binary logistic regression model. Will you select models with slightly worse performance if it makes more sense or is more parsimonious? Discuss why you selected your models.   -->

<!-- For the binary logistic regression model, will you use a metric such as log likelihood, AIC, ROC curve, etc.? Using the training data set, evaluate the binary logistic regression model based on (a) accuracy, (b) classification error rate, (c) precision, (d) sensitivity, (e) specificity, (f) F1 score, (g) AUC, and (h) confusion matrix. Make predictions using the evaluation data set. -->

####References

Model diagnostics: http://www.sthda.com/english/articles/36-classification-methods-essentials/148-logistic-regression-assumptions-and-diagnostics-in-r/

Leaps package usage: https://rstudio-pubs-static.s3.amazonaws.com/2897_9220b21cfc0c43a396ff9abf122bb351.html
  
####Appendix

#####load packages
library(knitr)
library(dplyr)
library(kableExtra)

#####load training data
url_crime_train <- 'https://raw.githubusercontent.com/littlejohnjeff/Data_621_Fall_2019/master/crime-training-data_modified.csv'
crime_train_data <- read.csv(url_crime_train, header = TRUE)
kable(crime_train_data[1:15,]) %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#####summarize training data
library(psych)
kable(psych::describe(crime_train_data)) %>% kable_styling(bootstrap_options = c("striped", "hover", "responsive"))

#####boxplots of each variable split by target value
library(reshape2)
library(dplyr)
library(ggplot2)
crime_plot <- melt(crime_train_data, id.vars= 'target') %>% mutate(target = as.factor(target))

ggplot(data = crime_plot, aes(x = variable, y = value)) + geom_boxplot(aes(fill = target)) + facet_wrap( ~ variable, , dir = "h", scales = 'free') + scale_fill_manual(values=c("blue3", "red3"))

#####histograms of training set
ggplot(crime_plot,aes(value)) + geom_histogram(bin=25,fill="Red3") + facet_wrap( ~ variable, , dir = "h", scales = 'free')

#####histrogram with log transformations of all variables
ggplot(crime_plot,aes(log(value))) + geom_histogram(bin=25,fill="Blue3") + facet_wrap( ~ variable, , dir = "h", scales = 'free')

#####correlations
library(stats)
library(corrplot)
cor_train <- cor(crime_train_data, method="pearson") 
kable(cor_train, "html") %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#####plot correlations
corrplot(cor_train)

#####correlations just with target value
cor_train_target <- as.data.frame(cor_train) %>% dplyr::select(target)
kable(cor_train_target, "html") %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#####lstat distribution
ggplot(crime_train_data,aes(x=lstat)) + geom_histogram(bin=25,fill="Red3") 

#####lstat distribution after log transformation
ggplot(crime_train_data,aes(x=log(lstat))) + geom_histogram(bin=25,fill="Blue3") 

#####split training data into true training and validation/tuning
train_size <- floor(0.8 * nrow(crime_train_data))

set.seed(123)
train_ind <- sample(seq_len(nrow(crime_train_data)), size = train_size)

crime_train_data_train <- crime_train_data[train_ind, ]
crime_train_data_valid <- crime_train_data[-train_ind, ]

#####build model 1 - all variables untransformed
model1_untransformed <- glm(formula = target ~ ., family = "binomial", data = crime_train_data_train)
summary(model1_untransformed)

#####model 1 logit relationships
library(tidyr)
probabilities <- predict(model1_untransformed, type = "response")
predicted.classes <- ifelse(probabilities > 0.5, "pos", "neg")
head(predicted.classes)

mydata <- crime_train_data_train %>%
  dplyr::select_if(is.numeric) 
predictors <- colnames(mydata)

mydata <- mydata %>%
  mutate(logit = log(probabilities/(1-probabilities))) %>%
  gather(key = "predictors", value = "predictor.value", -logit)

ggplot(mydata, aes(logit, predictor.value))+
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  theme_bw() + 
  facet_wrap(~predictors, scales = "free_y")

#####Cook's distance
#####https://stat.ethz.ch/R-manual/R-devel/library/stats/html/plot.lm.html
plot(model1_untransformed, which = 4, id.n = 3)

#####influential outlier checks
library(broom)
model1_untransformed.data <- augment(model1_untransformed) %>% 
  mutate(index = 1:n()) 

kable(model1_untransformed.data %>% top_n(3, .cooksd))  %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

#####graph standardized residuals
ggplot(model1_untransformed.data, aes(index, .std.resid)) + 
  geom_point(alpha = .5) +
  theme_light()

#####model 1 vif
library(car)
car::vif(model1_untransformed)

#####build model 2
model2_coll <- glm(formula = target ~ zn + indus + chas + nox + age + dis + rad + tax + ptratio + lstat + medv, family = "binomial", data = crime_train_data_train)
summary(model2_coll)

#####build model 3
model3_some_logs <- glm(formula = target ~ zn + indus + chas + nox + rm + age + dis + rad + tax + ptratio + log(lstat) + log(medv), family = "binomial", data = crime_train_data_train)
summary(model3_some_logs)

#####build model 4
model4_sig_var <- glm(formula = target ~ nox + rad + dis + ptratio + medv + tax + age, family = "binomial", data = crime_train_data_train)
summary(model4_sig_var)

#####build model 5
model5_leap <- glm(formula = target ~ nox + rad + tax + medv + age + indus, family = "binomial", data = crime_train_data_train)
summary(model5_leap)

#####build model 6
model6_step <- glm(formula = target ~ nox + rad + medv + age, family = "binomial", data = crime_train_data_train)
summary(model6_step)

#####model 1 confusion matrix
library(caret)
model1_predict <- predict(model1_untransformed, newdata=crime_train_data_valid, type="response")
model1_predict_data <- ifelse(model1_predict > .5, 1, 0)
confusionMatrix(data=model1_predict_data, crime_train_data_valid$target, positive='1')

#####model 1 ROC
library('pROC')
plot(roc(crime_train_data_valid$target, model1_predict))

#####model 2 confusion matrix
model2_predict <- predict(model2_coll, newdata=crime_train_data_valid, type="response")
model2_predict_data <- ifelse(model2_predict > .5, 1, 0)
confusionMatrix(data=model2_predict_data, crime_train_data_valid$target, positive='1')

#####model 2 roc
plot(roc(crime_train_data_valid$target, model2_predict))

#####model 5 confusion matrix
model5_predict <- predict(model5_leap, newdata=crime_train_data_valid, type="response")
model5_predict_data <- ifelse(model5_predict > .5, 1, 0)
confusionMatrix(data=model5_predict_data, crime_train_data_valid$target, positive='1')

#####model 5 roc
plot(roc(crime_train_data_valid$target, model5_predict))

#####model 6 confusion matrix
model6_predict <- predict(model6_step, newdata=crime_train_data_valid, type="response")
model6_predict_data <- ifelse(model6_predict > .5, 1, 0)
confusionMatrix(data=model6_predict_data, crime_train_data_valid$target, positive='1')

#####model 6 roc
plot(roc(crime_train_data_valid$target, model6_predict))

#####model 2 eval data histogram
url_crime_eval <- 'https://raw.githubusercontent.com/littlejohnjeff/Data_621_Fall_2019/master/crime-evaluation-data_modified.csv'
crime_eval_data <- read.csv(url_crime_eval, header = TRUE)
model2_eval_data <- predict(model2_coll, newdata=crime_eval_data, type="response")
model2_eval <- ifelse(model2_eval_data > .5, 1, 0)
hist(model2_eval)
crime_eval_data_out <- cbind(crime_eval_data,model2_eval)

#####training data histogram
hist(crime_train_data$target)

#####write eval result file from model 2
write.csv(crime_eval_data_out,"DATA 621Data_621_Hw_3_Evaluation_Output.csv",row.names = FALSE)

#####frontward steps using LEAPS - didn't run in rmarkdown but ran fine in Ru Studio
library(leaps)
reg_subsets.out <-
    regsubsets(target ~ .,
               data = crime_train_data_train,
               nbest = 1,       ##### 1 best model for each number of predictors
               nvmax = NULL,    ##### NULL for no limit on number of variables
               force.in = NULL, force.out = NULL,
               method = "exhaustive")



library(car)
layout(matrix(1:2, ncol = 2))
########## Adjusted R2
res.legend <-
    subsets(reg_subsets.out, statistic="adjr2", legend = FALSE, min.size = 5, main = "Adjusted R^2")



summary.out$adjr2


summary.out$which[6,]


#####frontward setps using MASS package - didn't run in rmarkdown but ran fine in Ru Studio
library(MASS)
mod6a1 <- lm(target ~ ., data=crime_train_data_train) 
mod6a2 <- lm(target ~ 1, data=crime_train_data_train) 
mod6a <- stepAIC(mod4a2, direction="forward", scope = list(upper=mod6a1, lower=mod6a2)) 
summary(mod6a) 
